---
title: "Tool Calling UI"
output:
  html_document:
    theme:
      version: 5
      preset: shiny
    self_contained: false
vignette: >
  %\VignetteIndexEntry{Tool Calling UI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
#| message: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
```

```{=html}
<style>
.shiny-tool-card {
  margin-bottom: 1rem !important;
}
</style>
```

## Introduction

shinychat automatically displays rich UI for tool calls and results, providing users with an intuitive view of what tools are being executed and their outcomes. This tool UI works out-of-the-box with ellmer and requires no additional configuration to get started.

By default, tool calls render as a quiet, collapsed activity row that summarizes what's happening without crowding the conversation. Users can expand that row to see each individual call, and drill into any call to see the full request and result. The rest of this article walks through the default collapsed/expanded behavior and the ways you can customize it---titles and icons, per-call labels and previews, how calls are grouped, and the content shown inside an expanded result.

## Basic tool display

Let's start with a simple weather forecasting tool to demonstrate the default behavior:

```{r}
#| echo: true
library(shinychat)
library(ellmer)
library(weathR) # for forecasts via `point_tomorrow()`

get_weather_forecast <- tool(
  function(lat, lon) {
    point_tomorrow(lat, lon, short = FALSE)
  },
  name = "get_weather_forecast",
  description = "Get the weather forecast for a location.",
  arguments = list(
    lat = type_number("Latitude"),
    lon = type_number("Longitude")
  )
)
```

```{r}
# Shim point_tomorrow with caching to speed up rendering
point_tomorrow <- local({
  result <- NULL
  function(lat, lon, short = TRUE) {
    if (is.null(result)) {
      result <<- weathR::point_tomorrow(lat, lon, short = short)
    }
    result
  }
})
```

With `{ellmer}`, you register this tool with a chat object.
When the LLM calls the tool, ellmer automatically evaluates the tool call and returns the result to the LLM.

```{r}
#| echo: true
#| eval: false

chat <- ellmer::chat("openai/gpt-4.1-nano", echo = "output")
chat$register_tool(get_weather_forecast)
chat$chat("What's the weather in Boston like today?")
```
```
◯ [tool call] get_weather_forecast(lat = 42.3601, lon = -71.0589)
● #> [{"time":"2025-08-05 11:00:00 EDT","temp":76,"dewpoint":17.7778,"hum…

The weather in Boston today is partly sunny with temperatures around 75-76°F
during the afternoon. There is some humidity at about 64-66%. There is a
possibility of rain starting around 4 PM, with increasing chances in the
evening. Winds are coming from the northeast at about 8-10 mph.
```

If you're interested in learning more about how to use and create tools with `{ellmer}`, we recommend reading the [Tool/function calling article](https://ellmer.tidyverse.org/articles/tool-calling.html) on the `{ellmer}` website.

If you're working interactively in the R console---and if you've set the `echo = "output"` option---ellmer shows you when tool calls are made and gives a preview of the tool result.

```
◯ [tool call] get_weather_forecast(lat = 42.3601, lon = -71.0589)
● #> [{"time":"2025-08-05 11:00:00 EDT","temp":76,"dewpoint":17.7778,"hum…
```

shinychat's tool UI works in the same way, but with rich output displays that are shown directly in the `chat_ui()` interface in Shiny apps.
When the LLM calls a tool, shinychat automatically displays the tool request in a collapsed card.
Expanding the card shows the arguments passed to the tool.

```{r}
#| echo: false

# Simulate a tool request
request <- ContentToolRequest(
  id = "tool_call_001",
  name = "get_weather_forecast",
  arguments = list(
    lat = 40.7128,
    lon = -74.0060
  ),
  tool = get_weather_forecast
)

contents_shinychat(request)
```

When the tool result is completed, shinychat replaces the request with a card containing the tool result:

```{r}
#| echo: false

request@id <- "tool_call_002" # Don't match so we don't hide the request

result <- ContentToolResult(
  value = get_weather_forecast(42.3601, -71.0589),
  request = request
)

contents_shinychat(result)
```

If the tool throws an error when called, ellmer captures the error and shows it to the LLM.
shinychat updates the tool card to show the error message:

```{r}
request@id <- "tool_call_001c"
error_result <- ContentToolResult(
  error = rlang::catch_cnd(mean[[1]]),
  request = request
)

contents_shinychat(error_result)
```

When you use `chat_app()` or `chat_server()`, shinychat automatically handles tool requests and results, displaying them in the chat interface.

On the other hand, if you're using `chat_ui()` and calling `chat_append()` to stream the chat output, you'll need to make sure that ellmer streams tool requests and results to shinychat by setting `stream = "content"` in the `$stream_async()` call.

```{r}
#| echo: true
#| eval: false
server <- function(input, output, session) {
  client <- ellmer::chat("openai/gpt-4.1-nano")
  client$register_tool(get_weather_forecast)

  observeEvent(input$chat_user_input, {
    stream <- client$stream_async(input$chat_user_input, stream = "content")
    chat_append("chat", stream)
  })
}
```

Setting `stream = "content"` tells ellmer to stream `ellmer::Content` objects rather than plain text.
As a result, `ellmer::ContentToolRequest` and `ellmer::ContentToolResult` objects are streamed to shinychat, which automatically displays the tool requests and results in the chat interface.

## Tool title and icon

### Via tool annotations

You can improve the visual presentation by adding `annotations` to your `ellmer::tool()` definition using `ellmer::tool_annotations()`.
If the tool annotations include a `title` or `icon`, shinychat will use them in the tool request and result displays.

```{r}
#| echo: true

get_weather_forecast <- tool(
  function(lat, lon) {
    point_tomorrow(lat, lon, short = FALSE)
  },
  name = "get_weather_forecast",
  description = "Get the weather forecast for a location.",
  arguments = list(
    lat = type_number("Latitude"),
    lon = type_number("Longitude")
  ),
  annotations = tool_annotations(
    title = "Weather Forecast",
    icon = bsicons::bs_icon("cloud-sun")
  )
)
```

Now the tool request shows the custom title

```{r}
request_with_annotations <- ContentToolRequest(
  id = "tool_call_003",
  name = "get_weather_forecast",
  arguments = list(
    lat = 42.3601,
    lon = -71.0589
  ),
  tool = get_weather_forecast
)

contents_shinychat(request_with_annotations)
```

and the tool results show both the custom title and the icon.

```{r}
request_with_annotations@id <- "tool_call_004"

result_with_annotations <- ContentToolResult(
  value = get_weather_forecast(42.3601, -71.0589),
  request = request_with_annotations
)

contents_shinychat(result_with_annotations)
```

### Present tense while running, past tense once done

A tool call's title can come from two independent places:

- The tool's **definition title**---set via `tool_annotations(title = )`---is shown while the call is running. Write it in the present tense, e.g. `"Running R code"`.
- The **result title**---set via `tool_result_display(title = )` (see [Customizing tool result display](#customizing-tool-result-display))---is shown once the first result for that call arrives. Write it in the past tense, e.g. `"Ran R code"`.

shinychat swaps from the definition title to the result title exactly once, the moment a result lands; it never rewrites or conjugates either string for you. If you don't supply a result title, the definition title stays put for the life of the card.

Older versions of shinychat wrapped the definition title in a client-side template---`"Running {title}"` while in progress, `"{title} failed"` on error. That template has been removed. If your tool's title reads a little oddly while running (e.g. it was written to read well only after the automatic "Running " prefix was added), add an explicit present-tense definition title, or a past-tense result title, to get the wording you want. Failures are no longer folded into the title string at all; they're shown as a separate status cue (a "failed" or "N failed" note with an icon) next to the title.

### Via the tool result

Using tool annotations is an easy way to set the title and icon for all tool requests and results, but sometimes you'll want to customize the display for specific tool calls or results.
In these cases, you'll need to update your tool function to return an `ellmer::ContentToolResult` object, which takes an `extra` property that includes a list of extra data to attach to the result.

shinychat looks for and uses a `display` object within `extra` to customize how the tool result is shown.
The recommended way to build this object is with `tool_result_display()`, which validates its arguments and documents every supported field; a bare named list with the same fields still works and is promoted internally, but `tool_result_display()` is preferred going forward.
If the display's `title` and `icon` are set, these values override the values in the tool's annotations.

One useful strategy is to include display parameters in the tool's function signature, allowing the LLM to include context or additional information in the tool call.
In the example below, we'll let the LLM write the tool result title and we'll pick an icon based on the forecasted temperatures.
Note that we need to also update the tool's `arguments` to include the new `location_name` parameter so that the LLM can provide a meaningful title.

```{r}
#| echo: true

get_weather_forecast <- tool(
  function(lat, lon, location_name) {
    forecast <- point_tomorrow(lat, lon, short = FALSE)

    icon <- if (any(forecast$temp > 70)) {
      bsicons::bs_icon("sun-fill")
    } else if (any(forecast$temp < 45)) {
      bsicons::bs_icon("snow")
    } else {
      bsicons::bs_icon("cloud-sun-fill")
    }

    ContentToolResult(
      forecast,
      extra = list(
        display = tool_result_display(
          title = paste("Weather Forecast for", location_name),
          icon = icon
        )
      )
    )
  },
  name = "get_weather_forecast",
  description = "Get the weather forecast for a location.",
  arguments = list(
    lat = type_number("Latitude"),
    lon = type_number("Longitude"),
    location_name = type_string("Name of the location for display to the user")
  ),
  annotations = tool_annotations(
    title = "Weather Forecast",
    icon = bsicons::bs_icon("cloud-sun")
  )
)
```

Now when the tool result is completed, shinychat shows the custom title and icon based on the forecasted temperature:

```{r}
request_title_icon_result <- ContentToolRequest(
  id = "tool_call_005",
  name = "get_weather_forecast",
  arguments = list(
    lat = 42.3601,
    lon = -71.0589,
    location_name = "Boston, MA"
  ),
  tool = get_weather_forecast
)

result_title_icon_result <- get_weather_forecast(42.3601, -71.0589, "Boston, MA")
result_title_icon_result@request <- request_title_icon_result

contents_shinychat(result_title_icon_result)
```

## Tool intent

In the last example, we saw that we could include arguments in the tool function to let the LLM write some of the display text for us (e.g., the location name).

This strategy is so useful that shinychat automatically supports it for an `` `_intent` `` argument.
When `` `_intent` `` is present in the tool's arguments, shinychat shows this value in the tool request and result titles as the _reason the tool was called_.

To demonstrate, we'll use take a break from our weather tool and create a tool that simply generates a random number.
On its own, the tool doesn't need any arguments, but we include a `` `_intent` `` argument in the tool function and we include a description to explain its purpose.

```{r}
#| echo: true

tool_random_number <- tool(
  function(`_intent`) {
    runif(1)
  },
  name = "tool_random_number",
  description = "Generate a random number.",
  arguments = list(
    `_intent` = type_string(
      "A short snippet used for display purposes to explain the call to the user."
    )
  ),
  annotations = tool_annotations(
    title = "Random Number Generator",
    icon = bsicons::bs_icon("dice-3-fill")
  )
)
```

When the tool is called, shinychat shows the reason the LLM called the tool in the header of the tool card.

```{r}
request_random_number <- ContentToolRequest(
  id = "tool_call_006",
  name = "tool_random_number",
  arguments = list(`_intent` = "Generate a random number for testing"),
  tool = tool_random_number
)

result_random_number <- ContentToolResult(
  runif(1),
  request = request_random_number
)

contents_shinychat(result_random_number)
```

## Customizing tool result display

We've already seen that we can [customize the tool title and icon](#via-the-tool-result) by returning a `ContentToolResult` object from our tool function and specifying `title` and `icon` in `extra$display`, built with `tool_result_display()`.

shinychat uses the `display` object for three additional categories of customization:

1. A short per-call `label` and `value_preview`, shown in the collapsed view.
2. Alternative `html`, `markdown`, or `text` to show the user in place of the text value shown to the LLM.
3. Options to control how the tool result is presented.

### Collapsed-view label and value preview

While a tool call is collapsed, shinychat shows a per-call `label`---a short identifying value like a filename or query, used to tell this call apart from other calls to the same tool---and a `value_preview`, a terse peek at the result, like `"1,204 rows"` or `"3 files updated"`. Both are optional and, like `title` and `icon`, are computed inside your tool function and returned via `tool_result_display()`:

```{r}
#| echo: true
#| eval: false

ContentToolResult(
  forecast_data,
  extra = list(
    display = tool_result_display(
      label = location_name,
      value_preview = paste(nrow(forecast_data), "hourly readings")
    )
  )
)
```

`label` and `value_preview` only affect the collapsed summary; the full result (or its [alternative display content](#alternative-display-content)) is still shown when the call is expanded.

### Alternative display content

By default, shinychat shows the tool result's `value` property as a code block to the user.
This is often sufficient, but in some cases your tool may collect or prepare data that could be better presented to users in a different format.

For example, our weather tool returns a data frame with the forecast for the next day.
The LLM sees a JSON representation of this data frame, but users would likely prefer to see a nicely formatted table.

shinychat looks for the following properties in the `display` list and shows the user the first available display in the following order:

1. `html`: An HTML string or object. This can be HTML generated via R packages like `{htmltools}`, `{gt}`, `{reactable}` or even `{htmlwidgets}`.
2. `markdown`: A markdown string that is automatically rendered as rich HTML in shinychat.
3. `text`: A plain text string that is shown without code formatting.

Keep in mind that the content you show to your users should faithfully represent the `value` shown to the LLM.
Our weather tool is a great example---a JSON object isn't very user-friendly, but a table showing the forecast data is perfect.

#### Alternative HTML display

In the following example, we'll update the `get_weather_forecast()` to show the user a nicely formatted HTML table using `{gt}`.

```{r}
#| echo: true

get_weather_forecast <- tool(
  function(lat, lon, location_name) {
    forecast_data <- point_tomorrow(lat, lon, short = FALSE)
    forecast_table <- gt::as_raw_html(gt::gt(forecast_data))

    ContentToolResult(
      forecast_data,
      extra = list(
        display = tool_result_display(
          html = forecast_table,
          title = paste("Weather Forecast for", location_name),
          full_screen = TRUE
        )
      )
    )
  },
  name = "get_weather_forecast",
  description = "Get the weather forecast for a location.",
  arguments = list(
    lat = type_number("Latitude"),
    lon = type_number("Longitude"),
    location_name = type_string("Name of the location for display to the user")
  ),
  annotations = tool_annotations(
    title = "Weather Forecast",
    icon = bsicons::bs_icon("cloud-sun")
  )
)
```

Expand the tool result card to see the forecast as a table.
Notice the fullscreen toggle button in the card body---clicking it expands the card to fill the viewport.

```{r}
request_html <- ContentToolRequest(
  id = "tool_call_007",
  name = "get_weather_forecast",
  arguments = list(location_name = "Boston, MA", lat = 42.3601, lon = -71.0589),
  tool = get_weather_forecast
)

result_html <- do.call(get_weather_forecast, request_html@arguments)
result_html@request <- request_html

contents_shinychat(result_html)
```

#### Alternative markdown display

You can also prepare markdown content to show your users based on the tool result.
It's less appropriate in our weather tool, but we could use markdown to summarize the forecast in a few sentences.


```{r}
#| echo: true

get_weather_forecast <- tool(
  function(lat, lon, location_name) {
    forecast_data <- point_tomorrow(lat, lon, short = FALSE)

    temp_current <- forecast_data$temp[1]
    skies_current <- forecast_data$skies[[1]]

    temp_high <- max(forecast_data$temp)
    temp_low <- min(forecast_data$temp)

    humidity <- round(mean(forecast_data$humidity), 1)
    skies <- table(forecast_data$skies)
    skies <- names(skies)[which.max(skies)]

    forecast_summary <- glue::glue(
      "In **{location_name}**, it's currently {temp_current}°F with _{tolower(skies_current)}_ skies. ",
      "Today's high will be {temp_high}°F and the low will be {temp_low}°F. ",
      "Humidity is around {humidity}%. ",
      "Look for **{tolower(skies)}** skies throughout the day."
    )

    ContentToolResult(
      forecast_data,
      extra = list(
        display = tool_result_display(
          markdown = forecast_summary,
          title = paste("Weather Forecast for", location_name)
        )
      )
    )
  },
  name = "get_weather_forecast",
  description = "Get the weather forecast for a location.",
  arguments = list(
    lat = type_number("Latitude"),
    lon = type_number("Longitude"),
    location_name = type_string("Name of the location for display to the user")
  ),
  annotations = tool_annotations(
    title = "Weather Forecast",
    icon = bsicons::bs_icon("cloud-sun")
  )
)
```

```{r}
request_html <- ContentToolRequest(
  id = "tool_call_008",
  name = "get_weather_forecast",
  arguments = list(location_name = "Boston, MA", lat = 42.3601, lon = -71.0589),
  tool = get_weather_forecast
)

result_html <- do.call(get_weather_forecast, request_html@arguments)
result_html@request <- request_html

contents_shinychat(result_html)
```


### Display Options

In addition to customizing the [tool title and icon](#via-the-tool-result) and providing alternative display content, you can also control how the tool result is presented using `tool_result_display()` arguments.

shinychat supports several options to control how results are presented:

- **`show_request = FALSE`**: Hide the tool call details from the tool results display.

   This is useful when you have rich output and the tool call details are clear from the rest of the display.
   For example, it might be appropriate to hide the request details when showing a [full formatted table of results](#alternative-html-display).

- **`open = TRUE`**: Expand the result panel by default.

   This is most useful when you've [customized the tool display](#alternative-display-content) to include an `{htmlwidget}` or other rich content that users should see immediately.

- **`full_screen = TRUE`**: Add a fullscreen toggle button to the tool result card.

   When clicked, the card expands to fill the entire viewport, making it easy to inspect large or detailed content like maps, tables, and plots. Users can exit fullscreen by pressing `Escape`, clicking the backdrop, or using the close button.

- **`title`** and **`icon`**: Choose the title and icon displayed in the tool result card header.

- **`label`** and **`value_preview`**: Choose the [per-call label and result preview](#collapsed-view-label-and-value-preview) shown in the collapsed view.


## Grouping tool calls

When a conversation includes several tool calls, shinychat can collapse them together into a single group rather than showing one card per call. Control this with `tool_grouping` on `chat_ui()`:

```{r}
#| echo: true
#| eval: false

chat_ui("chat", tool_grouping = "tool")
```

`tool_grouping` accepts one of:

- **`"tool"`** (the default): calls to the *same* tool made within one assistant turn's tool-calling phase are grouped together, regardless of the order in which they and other tools were called. For example, if a turn calls `get_weather_forecast`, then `search_web`, then `get_weather_forecast` again, the two `get_weather_forecast` calls are grouped even though a different tool call came between them.
- **`"all"`**: every tool call within the turn's tool-calling phase is grouped into a single collapsible summary, regardless of tool name.
- **`"none"`**: each tool call gets its own card; nothing is grouped.

Individual tools can override the chat-level setting with a `grouping` annotation:

```{r}
#| echo: true
#| eval: false

get_weather_forecast <- tool(
  function(lat, lon) {
    point_tomorrow(lat, lon, short = FALSE)
  },
  name = "get_weather_forecast",
  description = "Get the weather forecast for a location.",
  arguments = list(
    lat = type_number("Latitude"),
    lon = type_number("Longitude")
  ),
  annotations = tool_annotations(
    title = "Weather Forecast",
    grouping = "all"
  )
)
```

`grouping`, like `title` and `icon`, is a top-level field on the tool's annotations---no need to nest it under anything else.

## Global display options

shinychat uses the rich tool UI described above by default, but you can choose to hide all tool calls or use shinychat's basic display without allowing for rich or alternative content displays.

Most users won't need to customize these options, but it can be useful to use the basic display for debugging or verification purposes when you need to see the tool arguments and outputs exactly as written and seen by the LLM.

To adjust the display for debugging or verification purposes, set the `shinychat.tool_display` option (or the `SHINYCHAT_TOOL_DISPLAY` environment variable) to one of the following values:

```r
# Disable tool UI entirely
options(shinychat.tool_display = "none")

# Use basic text-based display (useful for verifying inputs/outputs)
options(shinychat.tool_display = "basic")

# Default rich display applies customizations from tool authors
options(shinychat.tool_display = "rich")
```

For most users, the default `"rich"` display provides the best experience.
